---
title: "GBM scRNA-seq: Iavarone functional state subtyping"
subtitle: "Mikolajewicz cohort · Singh Lab · v1.1"
author: "Singh Lab"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 8
    fig-height: 6
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r setup-knitr}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  fig.path = "figures/",
  dev = "png",
  dpi = 150
)
```

# Introduction

Glioblastoma (GBM) is the most common and aggressive primary brain tumour in adults. Malignant cells within a single tumour are not uniform: they sit in different **functional states** with distinct metabolic programs, neuronal mimicry, or proliferative capacity. Knowing which state a cell is in helps connect molecular targets to biology that is actually active in patient tissue.

Iavarone et al. defined four recurrent GBM tumour-cell states using spatial transcriptomics (CosMx) and validated them in other cohorts:

| State | Name | Biological theme |
|-------|------|------------------|
| **GPM** | Glycolytic / plurimetabolic | High glycolysis and flexible nutrient use |
| **MTC** | Mitochondrial | Oxidative phosphorylation and mitochondrial metabolism |
| **NEU** | Neuronal | Neuronal lineage mimicry / synaptic programs |
| **PPR** | Proliferative / progenitor | Cell cycle and DNA replication |

We do not reproduce their full single-cell GSEA pipeline here. Instead we **score each tumour cell against four published 13-gene signatures** and assign the state with the highest score (winner-take-all).

### Hypothesis

**PYGL** (muscle glycogen phosphorylase) and **PYGB** (brain glycogen phosphorylase) break down glycogen to glucose-1-phosphate, the first step in mobilising stored carbohydrate. The **GPM** state is defined by glycolytic and plurimetabolic programs, so we expect:

> **PYGL and PYGB expression should be enriched in GPM-classified tumour cells** compared with MTC, NEU, and PPR.

If that holds, glycogen metabolism (and PYGL/PYGB in particular) looks relevant in patient GBM, not just in cell lines.

### Data

- `data/gbm_seurat.rds`: processed Seurat object (Mikolajewicz snRNA-seq, ~18 patients)
- `data/iavarone_signatures.csv`: four gene signatures (Table S2 format)
- `data/mikolajewicz_metadata.csv`: optional per-cell metadata (same fields should be in `@meta.data`)

---

## Step 0: Setup

Load packages, record versions, set a random seed. UMAP and tie-breaking in `max.col` can depend on the seed, so we fix it for reproducibility.

```{r step0-setup}
library(Seurat)
library(tidyverse)
library(ggplot2)
library(patchwork)

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

use_ucell <- requireNamespace("UCell", quietly = TRUE)
if (!use_ucell) {
  message(
    "Package 'UCell' is not installed. Attempting Bioconductor install...\n",
    "If install fails, the notebook will fall back to Seurat::AddModuleScore."
  )
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }
  tryCatch(
    BiocManager::install("UCell", update = FALSE, ask = FALSE),
    error = function(e) message("UCell install failed: ", conditionMessage(e))
  )
  use_ucell <- requireNamespace("UCell", quietly = TRUE)
}
if (use_ucell) {
  library(UCell)
  scoring_method <- "UCell (rank-based; robust to sequencing depth)"
} else {
  scoring_method <- "AddModuleScore (mean-expression minus control genes)"
  message("Using Seurat::AddModuleScore as fallback.")
}

set.seed(42)

cat("Scoring method:", scoring_method, "\n\n")
print(sessionInfo())
```

`sessionInfo()` above has package versions. Worth keeping if you share or publish these results.

---

## Step 1: Load and inspect the Seurat object

The Seurat object holds expression **assays**, per-cell **meta.data** (patient, cell type, QC, etc.), and dimensionality **reductions** (PCA, UMAP). We print a summary before doing anything else to catch bad paths or missing columns early.

```{r step1-load}
seurat_path <- "data/gbm_seurat.rds"
stopifnot(file.exists(seurat_path))

obj <- readRDS(seurat_path)

cat("=== Seurat object summary ===\n")
print(obj)

cat("\n=== meta.data column names ===\n")
print(colnames(obj@meta.data))

cat("\n=== First rows of meta.data ===\n")
print(head(obj@meta.data))

cat("\n=== Assays ===\n")
print(Assays(obj))

cat("\n=== Reductions (embeddings) ===\n")
print(Reductions(obj))
```

### Metadata checks

We need `cell_type` (to subset tumour cells) and `sample` (patient ID for pseudobulk).

```{r step1-metadata-tables}
required_cols <- c("cell_type", "sample")
missing_cols <- setdiff(required_cols, colnames(obj@meta.data))
if (length(missing_cols) > 0) {
  stop(
    "Required metadata columns missing: ", paste(missing_cols, collapse = ", "),
    "\nAvailable: ", paste(colnames(obj@meta.data), collapse = ", ")
  )
}

cat("=== cell_type ===\n")
print(table(obj$cell_type, useNA = "ifany"))

cat("\n=== sample (cells per patient) ===\n")
print(sort(table(obj$sample), decreasing = TRUE))

cat("\n=== Unique patients ===\n")
print(length(unique(obj$sample)))
```

### What to expect

After the chunks above you should see roughly **10⁵ cells** across **18 patients** (`P1`–`P8`, `R1`–`R11`, `R9`). `cell_type` includes `tumor`, `neuron`, `myeloid`, `oligodendrocyte`, `astrocyte`, and others. The object usually has a normalised RNA assay (`SCT` or `RNA`) and may already include UMAP. If not, Step 6 can rebuild it from `umap_x` / `umap_y` in metadata.

---

## Step 2: Restrict to tumour cells

Iavarone's four states describe malignant tumour-cell programs. Neurons, myeloid, oligodendrocytes, and other non-tumour cells express overlapping genes (e.g. neuronal signature genes in real neurons) and would mess up state assignment if left in.

Subset to `cell_type == "tumor"` (American spelling in this object).

```{r step2-subset-tumor}
n_before <- ncol(obj)

if (!"tumor" %in% obj$cell_type) {
  stop(
    "'tumor' not found in cell_type. Unique values: ",
    paste(unique(obj$cell_type), collapse = ", ")
  )
}

obj_tumor <- subset(obj, subset = cell_type == "tumor")
n_after <- ncol(obj_tumor)

cat("Cells before tumour subset:", n_before, "\n")
cat("Cells after tumour subset: ", n_after, "\n")
cat("Fraction tumour: ", round(100 * n_after / n_before, 1), "%\n\n")

cat("=== Tumour cells per patient ===\n")
print(sort(table(obj_tumor$sample), decreasing = TRUE))
```

In many GBM snRNA-seq datasets tumour cells are a minority because the TME is cell-rich. In this Mikolajewicz cohort they are often the majority (~70% of cells), but it varies by patient. Either way, `n_after` is the denominator for all downstream state calls. Sparse patient × state combinations can add noise to pseudobulk later.

---

## Step 3: Load Iavarone signatures

The signature CSV has a slightly odd layout:

1. **Line 1**: title row (skip)
2. **Line 2**: column headers `GPM`, `MTC`, `NEU`, `PPR`
3. **Lines 3+**: ~13 genes per state; columns can differ in length

```{r step3-load-signatures}
sig_path <- "data/iavarone_signatures.csv"
stopifnot(file.exists(sig_path))

sig_df <- readr::read_csv(
  sig_path,
  skip = 1,
  col_names = TRUE,
  show_col_types = FALSE
)

signatures <- lapply(sig_df, function(col) {
  genes <- na.omit(col)
  genes <- genes[genes != ""]
  as.character(genes)
})
signatures <- signatures[intersect(names(signatures), c("GPM", "MTC", "NEU", "PPR"))]

cat("=== Signature gene counts ===\n")
for (nm in names(signatures)) {
  cat(sprintf("  %s: %d genes\n", nm, length(signatures[[nm]])))
}

obj_genes <- rownames(obj_tumor)

cat("\n=== Overlap with Seurat rownames ===\n")
sig_overlap <- lapply(signatures, function(genes) {
  present <- intersect(genes, obj_genes)
  missing <- setdiff(genes, obj_genes)
  list(present = present, missing = missing)
})

for (nm in names(sig_overlap)) {
  n_pres <- length(sig_overlap[[nm]]$present)
  n_tot  <- length(signatures[[nm]])
  cat(sprintf("  %s: %d / %d present in data\n", nm, n_pres, n_tot))
  if (length(sig_overlap[[nm]]$missing) > 0) {
    cat("    Missing:", paste(sig_overlap[[nm]]$missing, collapse = ", "), "\n")
  }
}
```

Gene symbols can differ across pipelines (Ensembl vs symbol, alias updates, dropout in snRNA-seq). UCell and AddModuleScore only use genes present in the matrix; missing genes are dropped silently. With only ~13 genes per signature, losing a few can shift scores noticeably. We print missing genes so you can judge how much to trust each state's score.

---

## Step 4: Score cells against the four signatures

For each cell and signature, module scoring asks whether that gene set is relatively highly expressed.

- **UCell** (`AddModuleScore_UCell`): rank-based; robust to library size.
- **AddModuleScore** (Seurat): mean signature expression minus matched control genes. Simpler but more sensitive to depth and normalisation.

Both give one numeric score per cell per signature. Higher = stronger evidence for that program.

```{r step4-score}
signatures_present <- lapply(signatures, function(g) intersect(g, obj_genes))

if (use_ucell) {
  obj_tumor <- UCell::AddModuleScore_UCell(
    obj_tumor,
    features = signatures_present,
    name = ""
  )
  score_cols <- names(signatures_present)
} else {
  obj_tumor <- Seurat::AddModuleScore(
    obj_tumor,
    features = signatures_present,
    name = "Iavarone_"
  )
  score_cols <- tail(
    colnames(obj_tumor@meta.data),
    length(signatures_present)
  )
  names_map <- setNames(score_cols, names(signatures_present))
  for (state in names(signatures_present)) {
    obj_tumor[[state]] <- obj_tumor@meta.data[[names_map[state]]]
  }
  score_cols <- names(signatures_present)
}

cat("Score columns used:", paste(score_cols, collapse = ", "), "\n\n")
cat("=== Summary of signature scores ===\n")
print(summary(obj_tumor@meta.data[, score_cols, drop = FALSE]))
```

Check the summaries for comparable scales. Big outliers might be ambient RNA or doublets.

---

## Step 5: Assign each cell to its top-scoring state

Winner-take-all: `iavarone_state` is whichever of GPM / MTC / NEU / PPR has the highest score (`max.col`).

```{r step5-assign-state}
score_mat <- as.matrix(obj_tumor@meta.data[, score_cols, drop = FALSE])

# ties go to the first max column (deterministic)
top_idx <- max.col(score_mat, ties.method = "first")
obj_tumor$iavarone_state <- factor(
  score_cols[top_idx],
  levels = c("GPM", "MTC", "NEU", "PPR")
)

cat("=== Cell counts per state ===\n")
state_tab <- table(obj_tumor$iavarone_state)
print(state_tab)

cat("\n=== Proportions ===\n")
print(round(prop.table(state_tab) * 100, 1))
```

Real cells can be mixed or transitional. One hard label throws away cells where two scores are close. For population-level hypothesis testing that is usually fine; for stratifying drug targets you might keep continuous scores instead.

---

## Step 6: UMAP

UMAP projects high-dimensional expression to 2D. Nearby points are transcriptionally similar. Clean colour separation suggests the signatures capture real variation; heavy mixing suggests gradual or ambiguous states.

```{r step6-umap}
if ("umap" %in% Reductions(obj_tumor)) {
  umap_source <- "Existing 'umap' reduction in Seurat object"
} else if (all(c("umap_x", "umap_y") %in% colnames(obj_tumor@meta.data))) {
  umap_source <- "Built from umap_x / umap_y columns in meta.data"
  umap_mat <- as.matrix(obj_tumor@meta.data[, c("umap_x", "umap_y")])
  rownames(umap_mat) <- colnames(obj_tumor)
  colnames(umap_mat) <- c("UMAP_1", "UMAP_2")
  obj_tumor[["umap"]] <- CreateDimReducObject(
    embeddings = umap_mat,
    key = "UMAP_",
    assay = DefaultAssay(obj_tumor)
  )
} else {
  umap_source <- "Recomputed (RunUMAP on PCA; slow on tumour-only subset)"
  obj_tumor <- NormalizeData(obj_tumor, verbose = FALSE)
  obj_tumor <- FindVariableFeatures(obj_tumor, verbose = FALSE)
  obj_tumor <- ScaleData(obj_tumor, verbose = FALSE)
  obj_tumor <- RunPCA(obj_tumor, verbose = FALSE)
  obj_tumor <- RunUMAP(obj_tumor, dims = 1:30, verbose = FALSE)
}

cat("UMAP source:", umap_source, "\n")

p_umap <- DimPlot(
  obj_tumor,
  reduction = "umap",
  group.by = "iavarone_state",
  cols = c(GPM = "#E41A1C", MTC = "#377EB8", NEU = "#4DAF4A", PPR = "#984EA3"),
  pt.size = 0.2
) +
  labs(
    title = "Iavarone functional states (tumour cells)",
    subtitle = umap_source
  ) +
  theme(plot.title = element_text(face = "bold"))

print(p_umap)

ggsave(
  "figures/umap_states.png",
  plot = p_umap,
  width = 8,
  height = 6,
  dpi = 150
)
```

Blobs of one colour = cells sharing that program. Intermingled colours = transitions or scoring ambiguity, which is common in GBM.

---

## Step 7: PYGL and PYGB across states

PYGL (liver/muscle isoform) and PYGB (brain isoform) catalyse glycogen breakdown. We test whether they track the **GPM** state.

Three checks, treating **PYGL** and **PYGB** separately:

1. **Pseudobulk plot**: mean expression per patient × state. Each point is one patient. Works better than single-cell violins when most cells are zero (dropout).
2. **Paired patient test**: within each patient, GPM mean vs pooled non-GPM; then test across patients.
3. **Spearman vs GPM score**: expression vs continuous GPM UCell score, no hard state labels.

We skip single-cell violins here. PYGL/PYGB are dropout-heavy and violins mostly show a spike at zero.

```{r step7-target-genes}
target_genes <- c("PYGL", "PYGB")
missing_targets <- setdiff(target_genes, obj_genes)

if (length(missing_targets) > 0) {
  warning(
    "TARGET GENE(S) MISSING FROM DATA: ",
    paste(missing_targets, collapse = ", "),
    "\nDownstream plots will skip missing genes."
  )
}
present_targets <- intersect(target_genes, obj_genes)
cat("Target genes present:", paste(present_targets, collapse = ", "), "\n")
```

### Pseudobulk expression

```{r step7-pseudobulk}
cat("DefaultAssay for expression:", DefaultAssay(obj_tumor), "\n")
expr_mat <- tryCatch(
  GetAssayData(obj_tumor, assay = DefaultAssay(obj_tumor), layer = "data"),
  error = function(e) GetAssayData(obj_tumor, assay = DefaultAssay(obj_tumor), slot = "data")
)

meta_pb <- obj_tumor@meta.data %>%
  tibble::rownames_to_column("cell") %>%
  dplyr::select(cell, sample, iavarone_state)

expr_long <- expr_mat[present_targets, , drop = FALSE] %>%
  as.matrix() %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("cell") %>%
  tidyr::pivot_longer(
    cols = all_of(present_targets),
    names_to = "gene",
    values_to = "expression"
  ) %>%
  dplyr::inner_join(meta_pb, by = "cell")

pseudobulk <- expr_long %>%
  dplyr::group_by(sample, iavarone_state, gene) %>%
  dplyr::summarise(
    mean_expression = mean(expression),
    n_cells = dplyr::n(),
    .groups = "drop"
  )

cat("=== Pseudobulk rows (patient × state × gene) ===\n")
print(head(pseudobulk, 12))

cat("\n=== Patients contributing per state (cell counts) ===\n")
print(
  pseudobulk %>%
    dplyr::group_by(iavarone_state) %>%
    dplyr::summarise(
      n_patients = dplyr::n_distinct(sample),
      total_cells = sum(n_cells),
      .groups = "drop"
    )
)
```

```{r step7-pseudobulk-plot}
if (nrow(pseudobulk) > 0) {
  state_fill <- c(GPM = "#E41A1C", MTC = "#377EB8", NEU = "#4DAF4A", PPR = "#984EA3")

  p_pb <- ggplot(pseudobulk, aes(x = iavarone_state, y = mean_expression, fill = iavarone_state)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.35, width = 0.55) +
    geom_point(aes(color = sample), position = position_jitter(width = 0.12), size = 2.8, alpha = 0.9) +
    facet_wrap(~ gene, scales = "free_y") +
    scale_fill_manual(values = state_fill) +
    labs(
      title = "PYGL and PYGB expression by Iavarone state (patient pseudobulk)",
      subtitle = "Each point = one patient; y = mean SCT-normalised expression within patient × state",
      x = "Iavarone state",
      y = "Mean expression (patient pseudobulk)"
    ) +
    theme_bw(base_size = 12) +
    theme(legend.position = "none", plot.title = element_text(face = "bold"))

  print(p_pb)

  ggsave(
    "figures/pseudobulk_PYGL_PYGB_by_state.png",
    plot = p_pb,
    width = 9,
    height = 5,
    dpi = 150
  )
}
```

### Quick numeric comparison (GPM vs others)

```{r step7-stats}
if (nrow(pseudobulk) > 0) {
  stats_summary <- pseudobulk %>%
    dplyr::group_by(gene, iavarone_state) %>%
    dplyr::summarise(
      mean_across_patients = mean(mean_expression),
      median_across_patients = median(mean_expression),
      n_patients = dplyr::n(),
      .groups = "drop"
    )

  cat("=== Pseudobulk mean expression per state (averaged across patients) ===\n")
  print(stats_summary)

  fc_tbl <- pseudobulk %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(
      n_gpm_obs = sum(iavarone_state == "GPM"),
      n_non_gpm_obs = sum(iavarone_state != "GPM"),
      mean_gpm = mean(mean_expression[iavarone_state == "GPM"], na.rm = TRUE),
      mean_non_gpm = mean(mean_expression[iavarone_state != "GPM"], na.rm = TRUE),
      fold_change_gpm_vs_rest = mean_gpm / mean_non_gpm,
      .groups = "drop"
    )

  cat("\n=== Fold change (GPM vs other states, pseudobulk) ===\n")
  print(fc_tbl)

  for (g in unique(pseudobulk$gene)) {
    sub <- pseudobulk %>% dplyr::filter(gene == g)
    if (dplyr::n_distinct(sub$iavarone_state) >= 2 && nrow(sub) >= 4) {
      kw <- kruskal.test(mean_expression ~ iavarone_state, data = sub)
      cat(sprintf("\nKruskal-Wallis for %s across states: p = %.4g\n", g, kw$p.value))
    }
  }
}
```

A t-test on 50k cells treats cells as independent and inflates significance. Patients (~18) are the real replicates for "is PYGL higher in GPM in GBM patients?"

### Paired test: GPM vs non-GPM within each patient

Each patient contributes a GPM mean and a non-GPM mean (MTC + NEU + PPR pooled). Pairing controls for between-patient differences in expression, purity, and treatment. With ~18 patients, n = 18 is the replicate count.

```{r step7-paired-test}
paired_stats <- NULL

if (exists("expr_long") && nrow(expr_long) > 0) {
  patient_paired <- expr_long %>%
    dplyr::group_by(sample, gene) %>%
    dplyr::summarise(
      gpm_mean = mean(expression[iavarone_state == "GPM"], na.rm = TRUE),
      rest_mean = mean(expression[iavarone_state != "GPM"], na.rm = TRUE),
      n_gpm = sum(iavarone_state == "GPM"),
      n_rest = sum(iavarone_state != "GPM"),
      .groups = "drop"
    ) %>%
    dplyr::filter(n_gpm > 0, n_rest > 0) %>%
    dplyr::mutate(paired_diff = gpm_mean - rest_mean)

  cat("=== Patients included per gene (must have GPM and non-GPM cells) ===\n")
  print(patient_paired %>% dplyr::count(gene, name = "n_patients"))

  paired_stats <- lapply(present_targets, function(g) {
    sub <- patient_paired %>% dplyr::filter(gene == g)
    if (nrow(sub) < 3) {
      return(tibble::tibble(
        gene = g, n_patients = nrow(sub), median_paired_diff = NA_real_,
        fold_change_gpm_vs_rest = NA_real_, wilcox_p = NA_real_, ttest_p = NA_real_,
        verdict = "Insufficient patients for testing"
      ))
    }
    wt <- wilcox.test(sub$gpm_mean, sub$rest_mean, paired = TRUE, exact = FALSE)
    tt <- t.test(sub$gpm_mean, sub$rest_mean, paired = TRUE)
    med_diff <- stats::median(sub$paired_diff)
    fc <- mean(sub$gpm_mean) / mean(sub$rest_mean)
    verdict <- if (!is.na(med_diff) && med_diff > 0 && wt$p.value < 0.05) {
      "Enrichment in GPM supported"
    } else {
      "Enrichment in GPM not supported"
    }
    tibble::tibble(
      gene = g,
      n_patients = nrow(sub),
      median_paired_diff = med_diff,
      fold_change_gpm_vs_rest = fc,
      wilcox_p = wt$p.value,
      ttest_p = tt$p.value,
      verdict = verdict
    )
  }) %>%
    dplyr::bind_rows()

  cat("\n=== Paired GPM vs non-GPM test results ===\n")
  print(paired_stats)

  for (i in seq_len(nrow(paired_stats))) {
    cat(sprintf(
      "\n%s: %s\n",
      paired_stats$gene[i], paired_stats$verdict[i]
    ))
  }
}
```

`median_paired_diff` > 0 means most patients have higher expression in GPM cells. Wilcoxon is the primary test; paired t-test is secondary.

### Spearman correlation with GPM score

Hard labels drop nuance. If PYGL tracks the GPM program, expression should correlate with the continuous **GPM UCell score** even before bucketing into states.

```{r step7-spearman}
spearman_stats <- NULL
gpm_col <- "GPM"
if (!gpm_col %in% colnames(obj_tumor@meta.data)) {
  stop("GPM score column not found in meta.data after module scoring.")
}

if (length(present_targets) > 0) {
  gpm_scores <- obj_tumor@meta.data[[gpm_col]]

  spearman_stats <- lapply(present_targets, function(g) {
    gene_expr <- as.numeric(expr_mat[g, ])
    ct <- stats::cor.test(gpm_scores, gene_expr, method = "spearman", exact = FALSE)
    interp_line <- if (ct$estimate >= 0.1 && ct$p.value < 0.05) {
      "Positive correlation with continuous GPM score"
    } else if (ct$estimate > 0 && ct$p.value < 0.05) {
      "Weak positive rho; significant at large n but small effect"
    } else if (ct$estimate > 0) {
      "Weak positive trend; not significant"
    } else {
      "No positive tracking of GPM score"
    }
    tibble::tibble(
      gene = g,
      spearman_rho = unname(ct$estimate),
      spearman_p = ct$p.value,
      interpretation = interp_line
    )
  }) %>%
    dplyr::bind_rows()

  cat("=== Spearman: GPM UCell score vs gene expression (all tumour cells) ===\n")
  print(spearman_stats)

  for (i in seq_len(nrow(spearman_stats))) {
    cat(sprintf(
      "\n%s: rho = %.3f, p = %.4g (%s)\n",
      spearman_stats$gene[i],
      spearman_stats$spearman_rho[i],
      spearman_stats$spearman_p[i],
      spearman_stats$interpretation[i]
    ))
  }
}
```

Positive Spearman rho means expression rises with GPM program strength. That can be more informative than categorical labels when states form gradients on the UMAP.

---

## Step 8: Interpretation

```{r step8-interpret-prep}
interp <- list(
  n_tumor = ncol(obj_tumor),
  n_patients = length(unique(obj_tumor$sample)),
  state_props = round(prop.table(table(obj_tumor$iavarone_state)) * 100, 1),
  scoring = scoring_method,
  umap = umap_source
)
if (exists("fc_tbl") && nrow(fc_tbl) > 0) interp$fc <- fc_tbl
if (exists("paired_stats") && nrow(paired_stats) > 0) interp$paired <- paired_stats
if (exists("spearman_stats") && nrow(spearman_stats) > 0) interp$spearman <- spearman_stats

fmt_paired <- function(g) {
  if (is.null(paired_stats) || !g %in% paired_stats$gene) return(paste0(g, " paired statistics not available."))
  row <- paired_stats[paired_stats$gene == g, ]
  paste0(
    "**Paired test (n = ", row$n_patients, " patients):** ",
    "median diff (GPM − rest) = ", round(row$median_paired_diff, 4),
    "; fold-change = ", round(row$fold_change_gpm_vs_rest, 2),
    "; Wilcoxon p = ", format.pval(row$wilcox_p, digits = 3),
    " (paired t-test p = ", format.pval(row$ttest_p, digits = 3), "). ",
    "**Verdict: ", row$verdict, ".**"
  )
}

fmt_spearman <- function(g) {
  if (is.null(spearman_stats) || !g %in% spearman_stats$gene) return("")
  row <- spearman_stats[spearman_stats$gene == g, ]
  paste0(
    "**Spearman vs GPM score:** rho = ", round(row$spearman_rho, 3),
    ", p = ", format.pval(row$spearman_p, digits = 3),
    ". ", row$interpretation, "."
  )
}

pygl_paired_line <- fmt_paired("PYGL")
pygb_paired_line <- fmt_paired("PYGB")
pygl_spearman_line <- fmt_spearman("PYGL")
pygb_spearman_line <- fmt_spearman("PYGB")

pygl_verdict <- if (!is.null(paired_stats) && "PYGL" %in% paired_stats$gene) {
  paired_stats$verdict[paired_stats$gene == "PYGL"]
} else NA_character_

pygb_verdict <- if (!is.null(paired_stats) && "PYGB" %in% paired_stats$gene) {
  paired_stats$verdict[paired_stats$gene == "PYGB"]
} else NA_character_

pygl_narrative <- if (identical(pygl_verdict, "Enrichment in GPM supported")) {
  "PYGL is higher in GPM than non-GPM cells at the patient level, and expression tracks the continuous GPM score. Glycogen mobilisation via PYGL fits the GPM program in patient GBM."
} else {
  "PYGL does not reach significance for GPM enrichment in this cohort. Check effect direction and the pseudobulk plot before concluding."
}

pygb_narrative <- if (identical(pygb_verdict, "Enrichment in GPM supported")) {
  "PYGB is enriched in GPM at the patient level."
} else {
  "PYGB trends toward GPM in pseudobulk but does not pass the paired patient test. Do not treat it the same as PYGL when discussing targets."
}

cat("=== Key numbers for interpretation ===\n")
str(interp, max.level = 2)
```

### Summary

Classified **`r ncol(obj_tumor)` tumour cells** from **`r length(unique(obj_tumor$sample))` patients** using **`r scoring_method`**.

**State distribution** (winner-take-all):

- GPM: `r state_tab["GPM"]` cells (`r interp$state_props["GPM"]`%)
- MTC: `r state_tab["MTC"]` cells (`r interp$state_props["MTC"]`%)
- NEU: `r state_tab["NEU"]` cells (`r interp$state_props["NEU"]`%)
- PPR: `r state_tab["PPR"]` cells (`r interp$state_props["PPR"]`%)

### UMAP

On `figures/umap_states.png`, GPM, NEU, and MTC often intermingle while PPR tends to separate. That matches other GBM work (e.g. Neftel et al.): proliferation is a discrete cell-cycle program, while metabolic and neuronal-mimicry states look more like gradients. Winner-take-all labels oversimplify those states, which is why we also ran Spearman on the continuous GPM score.

### PYGL

**Prediction:** PYGL should be enriched in GPM.

`r pygl_paired_line`

`r pygl_spearman_line`

`r pygl_narrative`

### PYGB

**Prediction:** PYGB should also be enriched in GPM.

`r pygb_paired_line`

`r pygb_spearman_line`

`r pygb_narrative`

### Caveats

| Issue | Why it matters |
|-------|----------------|
| **Small signatures (~13 genes)** | PPR missing H2AZ1/H4C3; a few dropouts shift scores. |
| **Winner-take-all** | Metabolic states are continuous; labels oversimplify (see UMAP). |
| **n ≈ 18 patients** | Limited power; report effect sizes, not just p-values. |
| **Uneven cells per patient × state** | Some pseudobulk / paired means average few cells. |
| **snRNA-seq dropout** | Glycogen genes are often detected in a minority of nuclei. |

### Takeaways

- **PYGL:** `r if (identical(pygl_verdict, "Enrichment in GPM supported")) "Paired testing and GPM score correlation support enrichment in GPM." else "No clear patient-level enrichment in GPM for PYGL in this cohort."`
- **PYGB:** `r if (identical(pygb_verdict, "Enrichment in GPM supported")) "Paired testing supports GPM enrichment for PYGB." else "No clear state enrichment for PYGB; keep PYGL and PYGB as separate hypotheses."`
- **UMAP:** PPR separation with mixed metabolic states is expected. Use continuous GPM scores alongside labels when prioritising targets.

---

## Session info

```{r session-info-final}
sessionInfo()
```
