---
title: "GBM scRNA-seq: Iavarone Subtyping (External Validation)"
subtitle: "Abdelfattah cohort, GSE182109 / Broad SCP1985"
author: "David George"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 8
    fig-height: 6
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r setup-knitr}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  fig.path = "figures/",
  dev = "png",
  dpi = 150
)
```

# Introduction

External validation of the Mikolajewicz Iavarone subtyping analysis on the Abdelfattah et al. GBM scRNA-seq cohort ([GSE182109](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE182109), [SCP1985](https://singlecell.broadinstitute.org/single_cell/study/SCP1985)).

Primary question: does PYGL enrichment in GPM tumour cells replicate at the patient level?

PYGB is a separate secondary hypothesis and is not pooled with PYGL.

The cohort mixes primary and recurrent Grade IV GBM (`Type` column). We report that mix as a limitation; it also asks whether any GPM/PYGL signal holds across disease settings.

### Data (`data/`)

| File | Role |
|------|------|
| `Processed_matrix.mtx.gz` + genes/barcodes tsV | Expression matrix |
| `Meta_GBM.txt.gz` | Cell metadata (SCP export) |
| `Cluster_GBM.txt.gz` | UMAP coordinates (X, Y) |
| `iavarone_signatures.csv` | GPM / MTC / NEU / PPR signatures |

### Mikolajewicz reference (v1.1)

| Gene | Paired Wilcoxon p | Fold-change (GPM vs rest) | Spearman rho (GPM score) | Verdict |
|------|-------------------|---------------------------|------------------------|---------|
| **PYGL** | 0.037 | 1.41 | 0.105 | GPM enrichment supported |
| **PYGB** | 0.163 | 1.12 | 0.051 | GPM enrichment not supported |

Abdelfattah results are compared to this table in Step 5.

---

## Step 0: Setup

```{r step0-setup}
library(Seurat)
library(Matrix)
library(tidyverse)
library(ggplot2)
library(patchwork)

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

use_ucell <- requireNamespace("UCell", quietly = TRUE)
if (!use_ucell) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
  BiocManager::install("UCell", update = FALSE, ask = FALSE)
  use_ucell <- requireNamespace("UCell", quietly = TRUE)
}
if (use_ucell) library(UCell)

tryCatch(utils::memory.limit(size = 24000), error = function(e) NULL)
if (exists("mem.maxVSize")) mem.maxVSize(24000)

set.seed(42)

# SCP files carry a TYPE annotation row on line 2; drop it after read
read_scp_table <- function(path) {
  df <- read.csv(gzfile(path), stringsAsFactors = FALSE, check.names = FALSE)
  if (nrow(df) > 0 && identical(df$NAME[1], "TYPE")) {
    df <- df[-1, , drop = FALSE]
  }
  df[!is.na(df$NAME) & df$NAME != "", , drop = FALSE]
}

cat("UCell available:", use_ucell, "\n\n")
print(sessionInfo())
```

---

## Step 1: Load and inspect

SCP exports use two header rows (column names, then `TYPE,group,...`). We read with standard headers and drop the TYPE row. Cell key: `NAME`.

```{r step1-load-mtx}
data_dir <- "data"
mtx_path <- file.path(data_dir, "Processed_matrix.mtx.gz")
genes_path <- file.path(data_dir, "Processed_genes.tsv.gz")
barcodes_path <- file.path(data_dir, "Processed_barcodes.tsv.gz")
meta_path <- file.path(data_dir, "Meta_GBM.txt.gz")
cluster_path <- file.path(data_dir, "Cluster_GBM.txt.gz")

for (p in c(mtx_path, genes_path, barcodes_path, meta_path, cluster_path, "data/iavarone_signatures.csv")) {
  stopifnot(file.exists(p))
}

counts <- ReadMtx(
  mtx = mtx_path,
  cells = barcodes_path,
  features = genes_path,
  feature.column = 1,
  cell.column = 1
)

cat("Matrix dimensions:", nrow(counts), "genes x", ncol(counts), "cells\n")
cat("PYGL in rownames:", "PYGL" %in% rownames(counts), "\n")
cat("PYGB in rownames:", "PYGB" %in% rownames(counts), "\n")

obj <- CreateSeuratObject(counts = counts, assay = "RNA", project = "Abdelfattah")
rm(counts); gc()

print(obj)
```

```{r step1-load-meta}
meta <- read_scp_table(meta_path)
rownames(meta) <- meta$NAME

matched <- sum(colnames(obj) %in% meta$NAME)
cat("Cells matching metadata:", matched, "/", ncol(obj), "\n")

if (matched < ncol(obj)) {
  obj <- obj[, colnames(obj) %in% meta$NAME]
}

obj <- AddMetaData(obj, meta[colnames(obj), , drop = FALSE])

print(table(obj$Assignment, useNA = "ifany"))
print(sort(table(obj$Patient), decreasing = TRUE))
print(table(obj$Assignment, obj$CopyKatPrediction))
print(table(obj$Type[obj$Assignment == "Glioma"], useNA = "ifany"))
print(table(obj$Grade[obj$Assignment == "Glioma"], useNA = "ifany"))
```

Among `Assignment == "Glioma"` cells, CopyKat calls ~67% aneuploid and ~29% diploid. We do not filter on CopyKat; `Assignment == "Glioma"` is the author malignant label.

```{r step1-load-umap}
clust <- read_scp_table(cluster_path)
clust <- clust[!is.na(suppressWarnings(as.numeric(clust$X))), , drop = FALSE]
clust$X <- as.numeric(clust$X)
clust$Y <- as.numeric(clust$Y)

umap_idx <- match(colnames(obj), clust$NAME)
if (any(is.na(umap_idx))) {
  obj <- obj[, !is.na(umap_idx)]
  umap_idx <- umap_idx[!is.na(umap_idx)]
}

umap_mat <- as.matrix(clust[umap_idx, c("X", "Y")])
rownames(umap_mat) <- colnames(obj)
colnames(umap_mat) <- c("UMAP_1", "UMAP_2")

obj[["umap"]] <- CreateDimReducObject(
  embeddings = umap_mat,
  key = "UMAP_",
  assay = DefaultAssay(obj)
)
```

---

## Step 2: Normalise and subset to Glioma

Log-normalise RNA (`NormalizeData`, `scale.factor = 10000`), then subset to `Assignment == "Glioma"`, matching Mikolajewicz `cell_type == "tumor"`.

```{r step2-prep}
obj <- NormalizeData(
  obj,
  assay = "RNA",
  normalization.method = "LogNormalize",
  scale.factor = 10000,
  verbose = FALSE
)

n_before <- ncol(obj)
obj_tumor <- subset(obj, subset = Assignment == "Glioma")
n_after <- ncol(obj_tumor)

rm(obj); gc()

cat("Glioma cells:", n_after, "of", n_before, "\n")
cat("Patients:", length(unique(obj_tumor$Patient)), "\n\n")
print(sort(table(obj_tumor$Patient), decreasing = TRUE))
print(table(obj_tumor$Type, useNA = "ifany"))
```

---

## Step 3: Iavarone scoring and states

```{r step3-signatures}
sig_df <- readr::read_csv("data/iavarone_signatures.csv", skip = 1, show_col_types = FALSE)
signatures <- lapply(sig_df, function(col) {
  genes <- na.omit(col)
  genes <- genes[genes != ""]
  as.character(genes)
})
signatures <- signatures[c("GPM", "MTC", "NEU", "PPR")]

obj_genes <- rownames(obj_tumor)
signatures_present <- lapply(signatures, function(g) intersect(g, obj_genes))
for (nm in names(signatures)) {
  miss <- setdiff(signatures[[nm]], signatures_present[[nm]])
  cat(sprintf("  %s: %d / %d present", nm, length(signatures_present[[nm]]), length(signatures[[nm]])))
  if (length(miss) > 0) cat(" | missing:", paste(miss, collapse = ", "))
  cat("\n")
}
```

```{r step3-score}
stopifnot(use_ucell)
obj_tumor <- UCell::AddModuleScore_UCell(
  obj_tumor,
  features = signatures_present,
  name = ""
)
score_cols <- names(signatures_present)

score_mat <- as.matrix(obj_tumor@meta.data[, score_cols, drop = FALSE])
obj_tumor$iavarone_state <- factor(
  score_cols[max.col(score_mat, ties.method = "first")],
  levels = c("GPM", "MTC", "NEU", "PPR")
)

state_tab <- table(obj_tumor$iavarone_state)
print(state_tab)
print(round(100 * prop.table(state_tab), 1))
```

NEU is ~61% of Glioma cells here vs ~43% in Mikolajewicz. State fractions are cohort-specific; replication is about PYGL within GPM, not matched state proportions.

```{r step3-umap}
state_cols <- c(GPM = "#E41A1C", MTC = "#377EB8", NEU = "#4DAF4A", PPR = "#984EA3")

p_umap <- DimPlot(
  obj_tumor,
  reduction = "umap",
  group.by = "iavarone_state",
  cols = state_cols,
  pt.size = 0.15
) +
  labs(
    title = "Iavarone states, Abdelfattah Glioma cells",
    subtitle = "UMAP from Cluster_GBM (not recomputed)"
  ) +
  theme(plot.title = element_text(face = "bold"))

print(p_umap)
ggsave("figures/umap_states_abdelfattah.png", p_umap, width = 8, height = 6, dpi = 150)
```

---

## Step 4: PYGL and PYGB

Pseudobulk plot (headline), paired patient test (GPM vs pooled non-GPM), cell-level Spearman vs GPM score. PYGL and PYGB analysed separately.

```{r step4-targets}
target_genes <- c("PYGL", "PYGB")
present_targets <- intersect(target_genes, rownames(obj_tumor))
stopifnot(length(present_targets) == 2)

expr_mat <- GetAssayData(obj_tumor, assay = "RNA", layer = "data")

expr_long <- expr_mat[present_targets, , drop = FALSE] %>%
  as.matrix() %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("cell") %>%
  tidyr::pivot_longer(all_of(present_targets), names_to = "gene", values_to = "expression") %>%
  dplyr::inner_join(
    obj_tumor@meta.data %>%
      tibble::rownames_to_column("cell") %>%
      dplyr::select(cell, Patient, iavarone_state),
    by = "cell"
  )

pseudobulk <- expr_long %>%
  dplyr::group_by(Patient, iavarone_state, gene) %>%
  dplyr::summarise(
    mean_expression = mean(expression),
    n_cells = dplyr::n(),
    .groups = "drop"
  )

print(head(pseudobulk, 8))
```

```{r step4-pseudobulk-plot}
p_pb <- ggplot(pseudobulk, aes(x = iavarone_state, y = mean_expression, fill = iavarone_state)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.35, width = 0.55) +
  geom_point(aes(color = Patient), position = position_jitter(width = 0.12), size = 2.5, alpha = 0.9) +
  facet_wrap(~ gene, scales = "free_y") +
  scale_fill_manual(values = state_cols) +
  labs(
    title = "PYGL / PYGB by Iavarone state (patient pseudobulk)",
    subtitle = "One point per patient",
    x = "Iavarone state",
    y = "Mean log-normalised expression"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none", plot.title = element_text(face = "bold"))

print(p_pb)
ggsave("figures/pseudobulk_abdelfattah.png", p_pb, width = 9, height = 5, dpi = 150)
```

### Paired patient-level test

```{r step4-paired}
glioma_per_patient <- obj_tumor@meta.data %>%
  dplyr::count(Patient, name = "n_glioma")

patient_paired <- expr_long %>%
  dplyr::group_by(Patient, gene) %>%
  dplyr::summarise(
    gpm_mean = mean(expression[iavarone_state == "GPM"], na.rm = TRUE),
    rest_mean = mean(expression[iavarone_state != "GPM"], na.rm = TRUE),
    n_gpm = sum(iavarone_state == "GPM"),
    n_rest = sum(iavarone_state != "GPM"),
    .groups = "drop"
  ) %>%
  dplyr::inner_join(glioma_per_patient, by = "Patient") %>%
  dplyr::filter(n_gpm > 0, n_rest > 0) %>%
  dplyr::mutate(paired_diff = gpm_mean - rest_mean)

run_paired_stats <- function(pp_df) {
  lapply(present_targets, function(g) {
    sub <- pp_df %>% dplyr::filter(gene == g)
    if (nrow(sub) < 3) {
      return(tibble::tibble(
        gene = g, n_patients = nrow(sub), median_paired_diff = NA_real_,
        fold_change_gpm_vs_rest = NA_real_, wilcox_p = NA_real_, ttest_p = NA_real_,
        n_patients_gpm_higher = NA_integer_, verdict = "Insufficient patients"
      ))
    }
    wt <- wilcox.test(sub$gpm_mean, sub$rest_mean, paired = TRUE, exact = FALSE)
    tt <- t.test(sub$gpm_mean, sub$rest_mean, paired = TRUE)
    tibble::tibble(
      gene = g,
      n_patients = nrow(sub),
      median_paired_diff = median(sub$paired_diff),
      fold_change_gpm_vs_rest = mean(sub$gpm_mean) / mean(sub$rest_mean),
      wilcox_p = wt$p.value,
      ttest_p = tt$p.value,
      n_patients_gpm_higher = sum(sub$paired_diff > 0),
      verdict = if (median(sub$paired_diff) > 0 && wt$p.value < 0.05) {
        "Enrichment in GPM supported"
      } else {
        "Enrichment in GPM not supported"
      }
    )
  }) %>% dplyr::bind_rows()
}

paired_stats <- run_paired_stats(patient_paired)
cat("=== All patients with GPM and non-GPM cells ===\n")
print(paired_stats)

# ndGBM-10 (1 cell), ndGBM-11 (21 cells): exclude patients with <50 Glioma cells
min_glioma_cells <- 50L
low_cell_patients <- glioma_per_patient %>%
  dplyr::filter(n_glioma < min_glioma_cells) %>%
  dplyr::pull(Patient)

cat("\nPatients with <", min_glioma_cells, " Glioma cells (excluded in sensitivity run):\n", sep = "")
print(glioma_per_patient %>% dplyr::filter(n_glioma < min_glioma_cells))

patient_paired_filt <- patient_paired %>%
  dplyr::filter(n_glioma >= min_glioma_cells)

paired_stats_filt <- run_paired_stats(patient_paired_filt)
paired_stats_filt$analysis <- "Excluding low-cell patients"
paired_stats$analysis <- "All eligible patients"

sensitivity_tbl <- dplyr::bind_rows(
  paired_stats %>% dplyr::select(analysis, dplyr::everything()),
  paired_stats_filt %>% dplyr::select(analysis, dplyr::everything())
)

cat("\n=== Sensitivity: paired Wilcoxon (full vs >=50 Glioma cells) ===\n")
print(sensitivity_tbl %>% dplyr::select(analysis, gene, n_patients, n_patients_gpm_higher,
                                        median_paired_diff, fold_change_gpm_vs_rest, wilcox_p, verdict))

pygl_verdict_unchanged <- identical(
  paired_stats$verdict[paired_stats$gene == "PYGL"],
  paired_stats_filt$verdict[paired_stats_filt$gene == "PYGL"]
)
cat("\nPYGL verdict unchanged after filtering:", pygl_verdict_unchanged, "\n")
```

### Spearman vs continuous GPM score

Cell-level Spearman is supplementary. Verdicts follow the paired patient test.

```{r step4-spearman}
gpm_scores <- obj_tumor@meta.data$GPM
n_cells <- ncol(obj_tumor)

spearman_stats <- lapply(present_targets, function(g) {
  gene_expr <- as.numeric(expr_mat[g, ])
  ct <- cor.test(gpm_scores, gene_expr, method = "spearman", exact = FALSE)
  paired_row <- paired_stats[paired_stats$gene == g, ]
  paired_supported <- identical(paired_row$verdict, "Enrichment in GPM supported")

  interp <- if (paired_supported) {
    if (ct$estimate >= 0.1) {
      "Cell-level rho is directionally consistent with the paired patient-level result."
    } else {
      "Modest cell-level rho; paired patient test is the primary readout."
    }
  } else {
    paste0(
      "Paired patient-level test is non-significant (primary). ",
      "Cell-level rho = ", round(unname(ct$estimate), 3),
      " is weak on its own; with ", format(n_cells, big.mark = ","),
      " cells, p is not interpretable as independent evidence."
    )
  }

  tibble::tibble(
    gene = g,
    spearman_rho = unname(ct$estimate),
    spearman_p = ct$p.value,
    interpretation = interp
  )
}) %>%
  dplyr::bind_rows()

print(spearman_stats)
```

---

## Step 5: Replication vs Mikolajewicz

```{r step5-interpret-prep}
mikol <- tibble::tribble(
  ~cohort, ~gene, ~wilcox_p, ~fold_change, ~spearman_rho, ~verdict,
  "Mikolajewicz", "PYGL", 0.037, 1.41, 0.105, "Enrichment in GPM supported",
  "Mikolajewicz", "PYGB", 0.163, 1.12, 0.051, "Enrichment in GPM not supported"
)

abdel <- paired_stats %>%
  dplyr::transmute(
    cohort = "Abdelfattah",
    gene,
    wilcox_p,
    fold_change = fold_change_gpm_vs_rest,
    spearman_rho = spearman_stats$spearman_rho[match(gene, spearman_stats$gene)],
    verdict
  )

replication_tbl <- dplyr::bind_rows(
  mikol %>% dplyr::select(cohort, gene, wilcox_p, fold_change, spearman_rho, verdict),
  abdel %>% dplyr::select(cohort, gene, wilcox_p, fold_change, spearman_rho, verdict)
)

print(replication_tbl)

fmt_paired <- function(g) {
  row <- paired_stats[paired_stats$gene == g, ]
  paste0(
    "Paired Wilcoxon (", row$n_patients, " patients; ",
    row$n_patients_gpm_higher, " of ", row$n_patients, " with GPM > rest): ",
    "median diff = ", round(row$median_paired_diff, 4),
    ", FC = ", round(row$fold_change_gpm_vs_rest, 2),
    ", p = ", format.pval(row$wilcox_p, digits = 3),
    ". Verdict: ", row$verdict, "."
  )
}

fmt_spearman <- function(g) {
  row <- spearman_stats[spearman_stats$gene == g, ]
  paste0("Cell-level Spearman rho = ", round(row$spearman_rho, 3), ". ", row$interpretation)
}

fmt_sensitivity <- function(g) {
  full <- paired_stats[paired_stats$gene == g, ]
  filt <- paired_stats_filt[paired_stats_filt$gene == g, ]
  unchanged <- identical(filt$verdict, full$verdict)
  paste0(
    "Sensitivity (>=50 Glioma cells): ",
    filt$n_patients_gpm_higher, " of ", filt$n_patients, " patients GPM > rest, ",
    "FC = ", round(filt$fold_change_gpm_vs_rest, 2), ", p = ", format.pval(filt$wilcox_p, digits = 3),
    "; verdict ", filt$verdict,
    if (unchanged) " (unchanged from full cohort)." else " (differs from full cohort)."
  )
}

pygl_paired_line <- fmt_paired("PYGL")
pygb_paired_line <- fmt_paired("PYGB")
pygl_spearman_line <- fmt_spearman("PYGL")
pygb_spearman_line <- fmt_spearman("PYGB")
pygl_sens_line <- fmt_sensitivity("PYGL")
pygb_sens_line <- fmt_sensitivity("PYGB")

pygl_replicates <- identical(
  paired_stats$verdict[paired_stats$gene == "PYGL"],
  "Enrichment in GPM supported"
)
pygb_replicates <- identical(
  paired_stats$verdict[paired_stats$gene == "PYGB"],
  "Enrichment in GPM not supported"
)

pygl_replication_text <- if (pygl_replicates) {
  paste0(
    "PYGL enrichment in GPM replicates in this cohort. ",
    "Wilcoxon p is smaller and fold-change is larger than Mikolajewicz (FC ~1.6 vs ~1.4), ",
    "but this is one additional cohort (n=17 patients in the paired test), not evidence of a universal effect."
  )
} else {
  "PYGL does not replicate the Mikolajewicz finding here; inspect direction and effect size."
}

pygb_replication_text <- if (pygb_replicates) {
  "The null PYGB result replicates: no significant GPM enrichment at the patient level in either cohort."
} else {
  "PYGB verdict differs between cohorts; inspect paired stats before concluding."
}
```

### PYGL

Mikolajewicz: p = 0.037, FC = 1.41, Spearman rho = 0.105, supported.

`r pygl_paired_line`

`r pygl_sens_line`

`r pygl_spearman_line`

Replication: `r pygl_replication_text`

### PYGB

Mikolajewicz: p = 0.16, FC = 1.12, Spearman rho = 0.051, not supported.

`r pygb_paired_line`

`r pygb_sens_line`

`r pygb_spearman_line`

Replication: `r pygb_replication_text`

### Limitations

We had 17 patients in the paired test (one lacked GPM or non-GPM cells). Two patients contributed very few Glioma cells (ndGBM-10: 1, ndGBM-11: 21); the sensitivity analysis excluding patients with fewer than 50 Glioma cells is reported above. Primary and recurrent tumours are mixed in `Type`. CopyKat called ~29% of Glioma cells diploid; we used author `Assignment` labels only. Non-GPM cells are NEU-heavy, so the pooled rest mean is not MTC-specific. PPR signatures may miss H2AZ1/H4C3 as in Mikolajewicz. States are assigned winner-take-all from continuous scores. Pseudobulk avoids single-cell dropout in the headline figure.

### Summary

PYGL is higher in GPM at the patient level in both cohorts; PYGB does not show clear GPM enrichment in either. Treat them as separate targets.

---

## Session info

```{r session-info}
sessionInfo()
```
